Arc Blanc: a real time ocean simulation framework

### 2025-01-20 paper.tex
Change notation of dx to delta to avoid confusion

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Finite-difference method}
\label{subsec:finiteDifferenceMethod}
Following Gomez (2000) \cite{gomezInteractiveSimulationWater2000},
\framework  uses the FDM with an explicit scheme to numerically solve the equation (\ref{eq:2dwave}).
It uses a square simulation zone $Z$ of dimension $L\times L$ around a given mesh $M$.
This zone is regularly discretized with a $dx$ step.
Time is also regularly discretized with a time step $dt$.

In the FDM simulation, \framework denotes by $h_{i,j}^n$ the water height $h(i\times dx,j\times dx,n \times dt)$ for discrete location $(i,j)\times dx$ and time $n\times dt$.
This leads to the following explicit scheme:
\begin{equation}
	\label{eq:fdm_scheme}
	h_{i,j}^{n+1} = a\left(h_{i+1,j}^n+h_{i-1,j}^n+h_{i,j+1}^n+h_{i,j-1}^n-4h_{i,j}^n\right) + 2h_{i,j}^n - h_{i,j}^{n-1}
\end{equation}
where
\begin{equation}
	a=\frac{c^2dt^2}{dx^2}.
\end{equation}
The Dirichlet boundary condition gives $h_{i,j}^n=0$ for $i$ or $j$ being on the boundary of $Z$.
Since a new wave is generated at each time step, a damping effect is added to make them disappear gradually.
Hence, the right part of Equation (\ref{eq:fdm_scheme}) is multiplied by the \textbf{damping factor} $d(t)=d^n$.
This damping factor is linearly related to the velocity of the mesh $v_M(t)$ to ensure numerical stability as follows:
\begin{equation}
	d(t) =
	\begin{cases}
		\left(1-\frac{||\mathbf{v_M}(t)||}{v_{max}}\right)d_0 + \frac{||\mathbf{v_M}(t)||}{v_{max}}d_{max} & \text{for~~}||\mathbf{v_M}(t)||\leq v_{max}, \\
		d_{max}                                                                          & \text{else}
	\end{cases}
\end{equation}
where $d_{max}$ is the maximum damping value obtained when the body's velocity is above a given threshold $v_{max}$,
and $d_0$ is the damping for a null velocity.
For a plausible appearance, \framework uses the following values by default:
\begin{equation}
	\begin{cases}
		d_0     & = 0.98,       \\
		d_{max} & = 0.999,      \\
		v_{max} & = 5 m.s^{-1}.
	\end{cases}
\end{equation}
Nevertheless, the user can adjust these values in a way that a body $M$ causes more or less interaction waves.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Grid translation}
\label{subsec:gridTranslation}
The waves generated by a given body $M$ are defined in its local coordinates system.
Hence, when $M$ moves, the waves have to move accordingly as done by Cords and Staadt (2009) \cite{cordsRealTimeOpenWater2009}.
A first solution consists of applying the translation to the grid at times $t$ and $t-dt$ and then of using the integration scheme of Equation (\ref{eq:fdm_scheme}).
Nevertheless, these time-consuming shifts can be avoided by using the following domain translation:
\begin{equation}
	\begin{cases}
		k = & i - \left\lfloor\frac{\mathbf{p}_x(t+dt)-\mathbf{p}_x(t)}{dx}\right\rfloor    \\
		l = & j - \left\lfloor\frac{\mathbf{p}_z(t+dt)-\mathbf{p}_z(t)}{dx}\right\rfloor    \\
		o = & i - \left\lfloor\frac{\mathbf{p}_x(t+dt)-\mathbf{p}_x(t-dt)}{dx}\right\rfloor \\
		p = & j - \left\lfloor\frac{\mathbf{p}_z(t+dt)-\mathbf{p}_z(t-dt)}{dx}\right\rfloor
	\end{cases}
\end{equation}
where $\mathbf p(t)=\left[p_x(t)\ p_z(t)\right]^T$ is the position of the mesh $M$ at time $t$, $k$ and $l$ (respectively $o$ and $p$) are the shifted indices defined by the translation of $M$ at times $t+dt$ from times $t$ (respectively from times $t-dt$).
Then, the integration scheme can be rewritten as follows:
\begin{equation}
	\label{eq:fdm_scheme_corrected}
	h_{i,j}^{n+1} = d^n\left(a\left(h_{k+1,l}^n+h_{k-1,l}^n+h_{k,l+1}^n+h_{k,l-1}^n- 4 h_{k,l}^n\right) + 2 h_{k,l}^n - h_{o,p}^{n-1}\right)
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Stability}
\label{subsec:stability}
The stability of the scheme defined by Equation (\ref{eq:fdm_scheme_corrected}) is related to the CFL condition (for Courant–Friedrichs–Lewy), here given as follows:
\begin{equation}
	\frac{c^2dt^2}{dx^2}\leq 0.5
\end{equation}
The parameter $dt$ is considered fixed throughout the simulation.
One contribution of this article is to adapt the simulation's parameters $c$ and $dx$ to the CFL, as presented below.

Preliminary tests have shown that $dx$ must be small for a low velocity $\mathbf{v_M}(t)$ of $M$.
Moreover, $dx$ must be large enough for $Z$ to encompass $M$ and the waves generated by $M$.
The size of $M$ is defined from its bounding box $(b_x, b_y, b_z)$ as $S(M)=\max(b_x, b_z)$.
\Framework lets the user define the bounds $dx_{min}$ and $dx_{max}$ so that $Z$ is at least twice as large as $M$.
Then $dx$ is defined as follows:
\begin{equation}
	dx =
	\begin{cases}
		0.999~dt                    & ~~ \text{if}~||\mathbf{v_M}(t)||<1    \\
		||\mathbf{v_M}(t)||0.999~dt & ~~ \text{if}~||\mathbf{v_M}(t)||\geq1
	\end{cases}
\end{equation}
Finally $dx$ is clamped between $dx_{min}$ and $dx_{max}$.
The value of $c$ is deduced from $dx$ and $dt$ to validate the CFL as follows:
\begin{equation}
	c = \sqrt{0.49} \frac{dx}{dt}
\end{equation}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Finite-difference method}
\label{subsec:finiteDifferenceMethod}
Following Gomez (2000) \cite{gomezInteractiveSimulationWater2000},
\framework  uses the FDM with an explicit scheme to numerically solve the equation (\ref{eq:2dwave}).
It uses a square simulation zone $Z$ of dimension $L\times L$ around a given mesh $M$.
This zone is regularly discretized with a $\delta$ step.
Time is also regularly discretized with a time step $dt$.

In the FDM simulation, \framework denotes by $h_{i,j}^n$ the water height $h(i\times \delta,j\times \delta,n \times dt)$ for discrete location $(i,j)\times \delta$ and time $n\times dt$.
This leads to the following explicit scheme:
\begin{equation}
	\label{eq:fdm_scheme}
	h_{i,j}^{n+1} = a\left(h_{i+1,j}^n+h_{i-1,j}^n+h_{i,j+1}^n+h_{i,j-1}^n-4h_{i,j}^n\right) + 2h_{i,j}^n - h_{i,j}^{n-1}
\end{equation}
where
\begin{equation}
	a=\frac{c^2dt^2}{\delta^2}.
\end{equation}
The Dirichlet boundary condition gives $h_{i,j}^n=0$ for $i$ or $j$ being on the boundary of $Z$.
Since a new wave is generated at each time step, a damping effect is added to make them disappear gradually.
Hence, the right part of Equation (\ref{eq:fdm_scheme}) is multiplied by the \textbf{damping factor} $d(t)=d^n$.
This damping factor is linearly related to the velocity of the mesh $v_M(t)$ to ensure numerical stability as follows:
\begin{equation}
	d(t) =
	\begin{cases}
		\left(1-\frac{||\mathbf{v_M}(t)||}{v_{max}}\right)d_0 + \frac{||\mathbf{v_M}(t)||}{v_{max}}d_{max} & \text{for~~}||\mathbf{v_M}(t)||\leq v_{max}, \\
		d_{max}                                                                          & \text{else}
	\end{cases}
\end{equation}
where $d_{max}$ is the maximum damping value obtained when the body's velocity is above a given threshold $v_{max}$,
and $d_0$ is the damping for a null velocity.
For a plausible appearance, \framework uses the following values by default:
\begin{equation}
	\begin{cases}
		d_0     & = 0.98,       \\
		d_{max} & = 0.999,      \\
		v_{max} & = 5 m.s^{-1}.
	\end{cases}
\end{equation}
Nevertheless, the user can adjust these values in a way that a body $M$ causes more or less interaction waves.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Grid translation}
\label{subsec:gridTranslation}
The waves generated by a given body $M$ are defined in its local coordinates system.
Hence, when $M$ moves, the waves have to move accordingly as done by Cords and Staadt (2009) \cite{cordsRealTimeOpenWater2009}.
A first solution consists of applying the translation to the grid at times $t$ and $t-dt$ and then of using the integration scheme of Equation (\ref{eq:fdm_scheme}).
Nevertheless, these time-consuming shifts can be avoided by using the following domain translation:
\begin{equation}
	\begin{cases}
		k = & i - \left\lfloor\frac{\mathbf{p}_x(t+dt)-\mathbf{p}_x(t)}{\delta}\right\rfloor    \\
		l = & j - \left\lfloor\frac{\mathbf{p}_z(t+dt)-\mathbf{p}_z(t)}{\delta}\right\rfloor    \\
		o = & i - \left\lfloor\frac{\mathbf{p}_x(t+dt)-\mathbf{p}_x(t-dt)}{\delta}\right\rfloor \\
		p = & j - \left\lfloor\frac{\mathbf{p}_z(t+dt)-\mathbf{p}_z(t-dt)}{\delta}\right\rfloor
	\end{cases}
\end{equation}
where $\mathbf p(t)=\left[p_x(t)\ p_z(t)\right]^T$ is the position of the mesh $M$ at time $t$, $k$ and $l$ (respectively $o$ and $p$) are the shifted indices defined by the translation of $M$ at times $t+dt$ from times $t$ (respectively from times $t-dt$).
Then, the integration scheme can be rewritten as follows:
\begin{equation}
	\label{eq:fdm_scheme_corrected}
	h_{i,j}^{n+1} = d^n\left(a\left(h_{k+1,l}^n+h_{k-1,l}^n+h_{k,l+1}^n+h_{k,l-1}^n- 4 h_{k,l}^n\right) + 2 h_{k,l}^n - h_{o,p}^{n-1}\right)
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Stability}
\label{subsec:stability}
The stability of the scheme defined by Equation (\ref{eq:fdm_scheme_corrected}) is related to the CFL condition (for Courant–Friedrichs–Lewy), here given as follows:
\begin{equation}
	\frac{c^2dt^2}{\delta^2}\leq 0.5
\end{equation}
The parameter $dt$ is considered fixed throughout the simulation.
One contribution of this article is to adapt the simulation's parameters $c$ and $\delta$ to the CFL, as presented below.

Preliminary tests have shown that $\delta$ must be small for a low velocity $\mathbf{v_M}(t)$ of $M$.
Moreover, $\delta$ must be large enough for $Z$ to encompass $M$ and the waves generated by $M$.
The size of $M$ is defined from its bounding box $(b_x, b_y, b_z)$ as $S(M)=\max(b_x, b_z)$.
\Framework lets the user define the bounds $\delta_{min}$ and $\delta_{max}$ so that $Z$ is at least twice as large as $M$.
Then $\delta$ is defined as follows:
\begin{equation}
	\delta =
	\begin{cases}
		0.999~dt                    & ~~ \text{if}~||\mathbf{v_M}(t)||<1    \\
		||\mathbf{v_M}(t)||0.999~dt & ~~ \text{if}~||\mathbf{v_M}(t)||\geq1
	\end{cases}
\end{equation}
Finally $\delta$ is clamped between $\delta_{min}$ and $\delta_{max}$.
The value of $c$ is deduced from $\delta$ and $dt$ to validate the CFL as follows:
\begin{equation}
	c = \sqrt{0.49} \frac{\delta}{dt}
\end{equation}
